---
title: "Classical Linear Regression Model & Ordinary Least Squares"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
    css: styles.css
execute:
  echo: true
  warning: false
  message: false
  error: true
---

::: callout-note
[Download PDF](week01.pdf)
:::

# Two variable linear regression analysis

## Readings

- Stock and Watson (2003), Chapters 4–5
- Dougherty (2016), Chapter 2
- Wooldridge (2013), Chapter 2

# Introduction

Econometrics means economic measurement. Econometricians attempt to quantify
economic relationships that are of interest to theory and policy.

# Correlation vs Regression

## Correlation

In economics we are interested in the relationship between two or more random
variables, for example:

- Sales and advertising expenditure
- Personal consumption and disposable income
- Investment and interest rates
- Earnings and schooling

A measure of linear association between two random variables $x$ and $y$
is the **covariance**, which for a sample of $n$ pairs of observations
$(x_1, y_1)$, $\ldots$, $(x_n, y_n)$ is calculated as

$$
\operatorname{cov}(x,y)
=
\frac{1}{n-1}
\sum_{i=1}^{n}
(x_i - \bar{x})(y_i - \bar{y})
$$

The covariance is a measure of **linear** association. It may be approximately
zero even when there is a strong non-linear (e.g. quadratic) relationship.

## Interactive Correlation Explorer

::: {.callout-tip}
## Explore Correlation Interactively!
Adjust the sliders below to see how different parameters affect the relationship between variables. This hands-on exploration helps build intuition for regression concepts.
:::

```{=html}
<div style="max-width: 1200px; margin: 0 auto; padding: 20px; background: linear-gradient(135deg, #667eea 0%, #764ba2 100%); border-radius: 15px; box-shadow: 0 10px 30px rgba(0,0,0,0.2);">
  
  <div style="background: white; padding: 25px; border-radius: 10px; margin-bottom: 20px;">
    
    <h3 style="margin-top: 0; color: #667eea; font-weight: 700;">🎮 Control Panel</h3>
    
    <div style="display: grid; grid-template-columns: repeat(auto-fit, minmax(250px, 1fr)); gap: 20px; margin-top: 20px;">
      
      <div>
        <label style="font-weight: 600; color: #333; display: block; margin-bottom: 8px;">📊 Correlation (ρ)</label>
        <input type="range" id="correlation" min="-1" max="1" step="0.1" value="0.7" style="width: 100%; height: 8px; border-radius: 5px;">
        <div id="corrValue" style="text-align: center; margin-top: 5px; font-weight: 700; color: #667eea; font-size: 18px;">0.7</div>
      </div>
      
      <div>
        <label style="font-weight: 600; color: #333; display: block; margin-bottom: 8px;">👥 Sample Size (n)</label>
        <input type="range" id="sampleSize" min="20" max="500" step="10" value="100" style="width: 100%;">
        <div id="sizeValue" style="text-align: center; margin-top: 5px; font-weight: 700; color: #667eea; font-size: 18px;">100</div>
      </div>
      
      <div>
        <label style="font-weight: 600; color: #333; display: block; margin-bottom: 8px;">🔊 Noise Level (σ)</label>
        <input type="range" id="noise" min="0.1" max="2" step="0.1" value="0.5" style="width: 100%;">
        <div id="noiseValue" style="text-align: center; margin-top: 5px; font-weight: 700; color: #667eea; font-size: 18px;">0.5</div>
      </div>
    </div>
    
    <div style="margin-top: 20px; display: flex; gap: 10px; flex-wrap: wrap;">
      <button id="regenerate" onclick="updatePlot()" style="padding: 12px 24px; background: linear-gradient(135deg, #667eea 0%, #764ba2 100%); color: white; border: none; border-radius: 8px; font-weight: 600; cursor: pointer; box-shadow: 0 4px 15px rgba(102, 126, 234, 0.4); transition: all 0.3s;">
        🎲 Regenerate Data
      </button>
      <button id="showStats" onclick="toggleStats()" style="padding: 12px 24px; background: linear-gradient(135deg, #f093fb 0%, #f5576c 100%); color: white; border: none; border-radius: 8px; font-weight: 600; cursor: pointer; box-shadow: 0 4px 15px rgba(245, 87, 108, 0.4); transition: all 0.3s;">
        📈 Show Regression Stats
      </button>
    </div>
  </div>
  
  <div id="plotContainer" style="background: white; padding: 20px; border-radius: 10px; box-shadow: 0 5px 20px rgba(0,0,0,0.1); min-height: 400px;"></div>
  
  <div id="statsContainer" style="background: white; padding: 20px; border-radius: 10px; margin-top: 20px; display: none;">
    <h4 style="color: #667eea; margin-top: 0;">📊 Regression Statistics</h4>
    <div id="statsContent"></div>
  </div>
</div>

<script src="https://cdn.plot.ly/plotly-latest.min.js"></script>
<script>
  let currentData = null;
  
  function randn() {
    let u = 0, v = 0;
    while(u === 0) u = Math.random();
    while(v === 0) v = Math.random();
    return Math.sqrt(-2.0 * Math.log(u)) * Math.cos(2.0 * Math.PI * v);
  }
  
  function generateData() {
    const n = parseInt(document.getElementById('sampleSize').value);
    const rho = parseFloat(document.getElementById('correlation').value);
    const sigma = parseFloat(document.getElementById('noise').value);
    
    const x = [];
    const y = [];
    
    for (let i = 0; i < n; i++) {
      const xi = randn();
      const noise = randn() * sigma;
      x.push(xi);
      y.push(rho * xi + noise);
    }
    
    return {x, y, n, rho, sigma};
  }
  
  function calculateRegression(x, y) {
    const n = x.length;
    const sumX = x.reduce((a, b) => a + b, 0);
    const sumY = y.reduce((a, b) => a + b, 0);
    const sumXY = x.reduce((sum, xi, i) => sum + xi * y[i], 0);
    const sumX2 = x.reduce((sum, xi) => sum + xi * xi, 0);
    
    const meanX = sumX / n;
    const meanY = sumY / n;
    
    const beta1 = (sumXY - n * meanX * meanY) / (sumX2 - n * meanX * meanX);
    const beta0 = meanY - beta1 * meanX;
    
    const yPred = x.map(xi => beta0 + beta1 * xi);
    const ssRes = y.reduce((sum, yi, i) => sum + Math.pow(yi - yPred[i], 2), 0);
    const ssTot = y.reduce((sum, yi) => sum + Math.pow(yi - meanY, 2), 0);
    const r2 = 1 - (ssRes / ssTot);
    
    const corrNumerator = x.reduce((sum, xi, i) => sum + (xi - meanX) * (y[i] - meanY), 0);
    const corrDenomX = Math.sqrt(x.reduce((sum, xi) => sum + Math.pow(xi - meanX, 2), 0));
    const corrDenomY = Math.sqrt(y.reduce((sum, yi) => sum + Math.pow(yi - meanY, 2), 0));
    const corr = corrNumerator / (corrDenomX * corrDenomY);
    
    return {beta0, beta1, r2, corr, yPred};
  }
  
  function updatePlot() {
    currentData = generateData();
    const {x, y, n, rho, sigma} = currentData;
    const {beta0, beta1, r2, corr, yPred} = calculateRegression(x, y);
    
    const trace1 = {
      x: x,
      y: y,
      mode: 'markers',
      type: 'scatter',
      name: 'Data Points',
      marker: {
        size: 10,
        color: '#667eea',
        opacity: 0.6,
        line: {color: 'white', width: 2}
      },
      hovertemplate: '<b>X:</b> %{x:.2f}<br><b>Y:</b> %{y:.2f}<extra></extra>'
    };
    
    const trace2 = {
      x: x,
      y: yPred,
      mode: 'lines',
      type: 'scatter',
      name: 'Regression Line',
      line: {color: '#f5576c', width: 3},
      hovertemplate: '<b>Predicted Y:</b> %{y:.2f}<extra></extra>'
    };
    
    const layout = {
      title: {
        text: `<b>Correlation Explorer</b><br><sub>ρ=${rho.toFixed(2)}, n=${n}, R²=${r2.toFixed(3)}</sub>`,
        font: {size: 20}
      },
      xaxis: {title: '<b>X Variable</b>', gridcolor: '#f0f0f0', zeroline: true, zerolinecolor: '#999'},
      yaxis: {title: '<b>Y Variable</b>', gridcolor: '#f0f0f0', zeroline: true, zerolinecolor: '#999'},
      plot_bgcolor: '#fafafa',
      paper_bgcolor: 'white',
      hovermode: 'closest',
      showlegend: true,
      legend: {x: 0.02, y: 0.98, bgcolor: 'rgba(255,255,255,0.9)', bordercolor: '#ddd', borderwidth: 1}
    };
    
    const config = {
      displayModeBar: true,
      displaylogo: false,
      modeBarButtonsToRemove: ['lasso2d', 'select2d']
    };
    
    Plotly.newPlot('plotContainer', [trace1, trace2], layout, config);
    updateStats(beta0, beta1, r2, corr, n);
  }
  
  function updateStats(beta0, beta1, r2, corr, n) {
    const statsHTML = `
      <div style='display: grid; grid-template-columns: repeat(auto-fit, minmax(200px, 1fr)); gap: 15px;'>
        <div style='padding: 15px; background: linear-gradient(135deg, #667eea 0%, #764ba2 100%); border-radius: 8px; color: white;'>
          <div style='font-size: 12px; opacity: 0.9;'>Intercept (β₀)</div>
          <div style='font-size: 24px; font-weight: 700;'>${beta0.toFixed(3)}</div>
        </div>
        <div style='padding: 15px; background: linear-gradient(135deg, #f093fb 0%, #f5576c 100%); border-radius: 8px; color: white;'>
          <div style='font-size: 12px; opacity: 0.9;'>Slope (β₁)</div>
          <div style='font-size: 24px; font-weight: 700;'>${beta1.toFixed(3)}</div>
        </div>
        <div style='padding: 15px; background: linear-gradient(135deg, #4facfe 0%, #00f2fe 100%); border-radius: 8px; color: white;'>
          <div style='font-size: 12px; opacity: 0.9;'>R-squared</div>
          <div style='font-size: 24px; font-weight: 700;'>${r2.toFixed(3)}</div>
        </div>
        <div style='padding: 15px; background: linear-gradient(135deg, #43e97b 0%, #38f9d7 100%); border-radius: 8px; color: white;'>
          <div style='font-size: 12px; opacity: 0.9;'>Correlation</div>
          <div style='font-size: 24px; font-weight: 700;'>${corr.toFixed(3)}</div>
        </div>
      </div>
      <div style='margin-top: 20px; padding: 15px; background: #f5f5f5; border-radius: 8px; border-left: 4px solid #667eea;'>
        <strong>Regression Equation:</strong> 
        <code style='background: white; padding: 5px 10px; border-radius: 4px; display: inline-block; margin-top: 5px;'>
          Y = ${beta0.toFixed(3)} + ${beta1.toFixed(3)}·X
        </code>
      </div>
    `;
    document.getElementById('statsContent').innerHTML = statsHTML;
  }
  
  function toggleStats() {
    const container = document.getElementById('statsContainer');
    container.style.display = container.style.display === 'none' ? 'block' : 'none';
  }
  
  document.getElementById('correlation').addEventListener('input', function(e) {
    document.getElementById('corrValue').textContent = e.target.value;
  });
  
  document.getElementById('sampleSize').addEventListener('input', function(e) {
    document.getElementById('sizeValue').textContent = e.target.value;
  });
  
  document.getElementById('noise').addEventListener('input', function(e) {
    document.getElementById('noiseValue').textContent = e.target.value;
  });
  
  if (document.readyState === 'loading') {
    document.addEventListener('DOMContentLoaded', updatePlot);
  } else {
    updatePlot();
  }
</script>
```

::: {.callout-note}
## 🎓 Learning Objectives
- **Experiment** with correlation strength to see how it affects scatter patterns
- **Observe** how sample size influences the precision of estimates
- **Understand** the role of noise (random error) in obscuring relationships
- **Compare** the actual correlation with the estimated R-squared value
:::

## Understanding Different Correlation Patterns

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

library(plotly)

set.seed(456)
scenarios <- list(
  list(name = "Strong Positive (r=0.9)", rho = 0.9, color = "#43e97b"),
  list(name = "Moderate Positive (r=0.5)", rho = 0.5, color = "#4facfe"),
  list(name = "No Correlation (r=0)", rho = 0, color = "#f5576c"),
  list(name = "Moderate Negative (r=-0.5)", rho = -0.5, color = "#f093fb"),
  list(name = "Strong Negative (r=-0.9)", rho = -0.9, color = "#764ba2")
)

fig <- plot_ly()

for (scenario in scenarios) {
  x <- rnorm(80)
  y <- scenario$rho * x + rnorm(80, sd = 0.4)
  
  fig <- fig %>%
    add_trace(
      x = x,
      y = y,
      type = 'scatter',
      mode = 'markers',
      name = scenario$name,
      marker = list(
        size = 8,
        color = scenario$color,
        opacity = 0.6,
        line = list(color = 'white', width = 1)
      ),
      hovertemplate = paste0(
        '<b>', scenario$name, '</b><br>',
        'X: %{x:.2f}<br>',
        'Y: %{y:.2f}<extra></extra>'
      )
    )
}

fig <- fig %>%
  layout(
    title = list(
      text = '<b>Comparing Different Correlation Strengths</b>',
      font = list(size = 18)
    ),
    xaxis = list(title = '<b>X Variable</b>', gridcolor = '#f0f0f0'),
    yaxis = list(title = '<b>Y Variable</b>', gridcolor = '#f0f0f0'),
    plot_bgcolor = '#fafafa',
    paper_bgcolor = 'white',
    hovermode = 'closest',
    legend = list(
      orientation = 'h',
      y = -0.15,
      x = 0.5,
      xanchor = 'center'
    )
  )

fig
```

## Overview

In this handout we will revisit the Classical Linear Regression Model (CLRM) [see @wooldridge2010, chap. 1-2]. The goal of this week's lecture is to:

1. understand the model specification;
2. its underlying assumptions;
3. and the appropriate interpretation;
4. the OLS estimator, using linear algebra;
5. the geometry of OLS and partitioned regression result.

## Model Specification

The linear population regression model is given by,

$$ 
\begin{aligned}
  Y_i =& X_i'\beta+\varepsilon_i \\
  =& \beta_1\mathbf{1}+\beta_2X_{i2}+\beta_3X_{i3}+...+\beta_kX_{ik}+\varepsilon_i
\end{aligned}
$$


# R Code Example

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false
#| error: true

library(tidyverse)
library(haven)

# Read the Stata dataset (adjust path if needed)
qlfs18 <- read_dta("data/qlfs1314_2.dta")

# Show variables so you can verify names during render
head(names(qlfs18), 30)


summary(qlfs18$lhourpay)

```




