---
title: "Classical Linear Regression Model & Ordinary Least Squares"

format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
  pdf:
    toc: true
    number-sections: true
    output-file: "week01.pdf"

execute:
  echo: true     # show code
  eval: true     # run code
  warning: true  # show warnings (change to false if you want)
  message: true  # show messages (change to false if you want)
  error: true    # IMPORTANT: don't stop rendering if one chunk errors
---

::: callout-note
[Download PDF](week01.pdf)
:::

# Two variable linear regression analysis

## Readings
- Stock and Watson (2003), Chapters 4–5
- Dougherty (2016), Chapter 2
- Wooldridge (2013), Chapter 2

# Introduction
Econometrics means economic measurement. Econometricians attempt to quantify
economic relationships that are of interest to theory and policy.


# Correlation vs Regression

## Correlation
In economics we are interested in the relationship between two or more random
variables, for example:
- Sales and advertising expenditure
- Personal consumption and disposable income
- Investment and interest rates
- Earnings and schooling
A measure of linear association between two random variables $x$ and $y$
is the **covariance**, which for a sample of $n$ pairs of observations
$(x_1, y_1)$, $\ldots$, $(x_n, y_n)$ is calculated as

$$
\operatorname{cov}(x,y)
=
\frac{1}{n-1}
\sum_{i=1}^{n}
(x_i - \bar{x})(y_i - \bar{y})
$$

The covariance is a measure of **linear** association. It may be approximately
zero even when there is a strong non-linear (e.g. quadratic) relationship.

# Overview
In this handout we will revisit the Classical Linear Regression Model (CLRM) [see @wooldridge2010, chap. 1-2]. The goal of this week's lecture is to:
1.  understand the model specification;
2.  it's underlying assumptions;
3.  and the appropriate interpretation;
4.  the OLS estimator, using linear algebra;
5.  the geometry of OLS and partitioned regression result.

# Model Specification
The linear population regression model is given by,

$$ 
\begin{aligned}
  Y_i =& X_i'\beta+\varepsilon_i \\
  =& \beta_1\mathbf{1}+\beta_2X_{i2}+\beta_3X_{i3}+...+\beta_kX_{ik}+\varepsilon_i
\end{aligned}
$$
```{r}
#| echo: true
#| eval: true

1+1

qlfs18 <- haven::read_dta("data/qlfs1314_2.dta")

```

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false
#| error: true

library(tidyverse)
library(haven)

# Read the Stata dataset (adjust path if needed)
qlfs18 <- read_dta("data/qlfs1314_2.dta")

# Show variables so you can verify names during render
head(names(qlfs18), 30)


summary(qlfs18$lhourpay)

```

